Squeezing in Bose-Einstein condensates with large numbers of atoms

### 2013-10-10 Response to referees.txt

We thank the referees for their detailed reports.  We have addressed their 
concerns and clarified answers to their questions in a new version of the 
manuscript.

Response to suggestions of referee 1:

1. We have clarified the wording of the discussion at the bottom of page 4, to
better explain the derivation of Eq. (12), which is the reduced, approximate 
form of (11).

2. We had indeed overlooked defining \tau. We have now added the definition
of \tau, and renamed the period of cyclic variations from \tau to T to
avoid confusion.

3. We have also added a paragraph at the end of Section 3 clearly explaining
our definition of number squeezing in the multimode case, and refer to this
definition in the caption of Figure 4.

4. We thank the referee for bring the extra reference on atomic Kerr squeezing
to our attention, and we have added it to our paper.

Response to questions and suggestions of referee 2:

1.  We have added a discussion of the practicality of using squeezing to improve a precision 
measurement relative to the possibility of increasing the atom number.  In fact, these two methods 
should ideally be used together if possible, as they are essentially entirely compatible and 
independent improvements to the sensitivity.  We include the highly relevant reference that the 
referee quoted.

2.  We have added a comment that the large number, box-potential geometries are most obviously
adapted to precision inertial measurements rather than in-trap measurements with goals of spatial 
resolution.

3.  We have rewritten our discussion of the effect of trap geometry on squeezing in response to 
the referee's very reasonable comments that it was confusing and apparently contradictory.  The 
effect of dimensionality changes character completely depending on whether the density is kept 
constant, or the total atom number is kept constant.  We have elaborated on this point, and made
the manuscript clear on this key issue.

     These two conclusions seem contradictory, and it is not clear how to relate them. This is
    made more difficult because figure 10 is missing the part labelled "Right", which as I
    understand should show the achievable squeezing in the 3D case. The authors appear to
    resolve this seeming contradiction in the conclusion, but it is unclear from the text how
    this relates to the parameters used in the different simulations.

### 2013-10-10 large_number_squeezing_with_atoms.tex
Nearly ready to resubmit

Producing squeezed states for metrological purposes is therefore only relevant when it can be achieved in a context of large numbers of atoms.  This paper examines the generation of atomic squeezing using the inherent Kerr-like nonlinearities of ultracold atoms, and demonstrates a trap geometry in which metrologically relevant levels of squeezing are present in trap geometries that are compatible with large atomic number.

@@

In the multimode case, the single mode definition for number squeezing, $\langle\mathrm{Var}[\hat{N}_i] \rangle / \langle \hat{N}_i \rangle$, where $\hat{N}_i$ is the number operator for state $|i\rangle \in \{|a\rangle, |b\rangle \}$, is no longer adequate. In the multimode case we define the number squeezing in states $|i\rangle$ to be
\begin{equation}
\mathrm{Squeezing} = \frac{\mathrm{Var} \left[ \sum_j \hat{N}_i(x_j) \right]} { \sum_j \hat{N}_i(x_j) }
\label{eqMultimodeSqueezingDefinition}
\end{equation}
where the sums are over spatial modes defined as delta functions in position, and there is seperate number operator $\hat{N}_i(x_j)$ for each mode $j$ in state $|i\rangle$.

@@

The simulations in the previous two sections were one dimensional, and show how having spatially constant density and using a $\pi$-pulse will improve the resultant squeezing.  While these conclusions remain valid for simulations in two or three dimensions, adding each additional dimension increases the number of spatial modes in a given energy range.  For BECs of different dimension at the same energy per particle, this leads to a degradation of squeezing, even when the potential (and the resulting atomic density) is constant across the bulk of the trap.  This is demonstrated in Fig.~\ref{figDimensionalSqueezingEffects}, where we show the squeezing generated in 1D, 2D and 3D BECs for two different parameter regimes. In both cases we take the BEC to be trapped in a box potential, which in the Thomas-Fermi limit leads to a constant density profile across the BEC. In Figure~\ref{figDimensionalSqueezingEffects}a we consider the case where $U_{aa}=U_{bb}\neq 0$, $U_{ab}=0$, while in Figure~\ref{figDimensionalSqueezingEffects}b we consider the case $U_{aa}\neq 0$, $U_{ab}=U_{bb}=0$. In both cases it is clear that number squeezing is best in lower dimensions, and becomes degraded or non-existent as we move to a three dimensional system; the effect is worse if more than one nonlinearity is non-zero.

@@

We wish to examine the scaling of the non-condensed fraction $N_{\mathbf{k}\neq0}/N$ with the physical size of the condensate.  Other system parameters such as the total number $N$ and the occupation of the $\ket{a}$ and $\ket{b}$ states at $t_1$ ($n_a$ and $n_b$, respectively) are kept constant and chosen to maximise the squeezing of the equivalent two-mode model presented in Section~\ref{secTwoModeAnalytic}.  In particular, this fixes the values of $\theta$, $\phi$, and $\lambda_{ij} = \chi_{ij}\tau_\text{hold}$.  As the size of the condensate is scaled, $\hbar \chi_{ij} = U^\text{1D}_{ij}/L = U^\text{2D}_{ij}/A = U^\text{3D}_{ij}/V$ will change, but we will vary $\tau_\text{hold}$ to keep $\lambda_{ij}$ constant.  Therefore $\Lambda_a = n_a\chi_{aa}\tau_\text{hold}$ will also be constant.  With this in mind, we can write $\tau_\text{hold}$ in terms of the system size and the nonlinearity $U_{aa}$.  The non-condensed fraction can then be shown to scale with the physical size of the condensate as:
\begin{align}
  \frac{N^\text{1D}_{\mathbf{k}\neq 0}(t_3)}{N} &\approx \sqrt{L} \cdot \frac{1}{N} \sqrt{\frac{M U^\text{1D}_{aa} n_a}{2 \hbar^2 \pi^2}} \Lambda_a^{3/2} \cos^2\theta f^\text{1D}(\Lambda_a) &&\propto \sqrt{L}, \\
  \frac{N^\text{2D}_{\mathbf{k}\neq 0}(t_3)}{N} &\approx 1 \cdot \frac{1}{N} \frac{M U^\text{2D}_{aa} n_a }{4 \hbar^2} \Lambda_a \cos^2\theta f^\text{2D}(\Lambda_a) &&\propto \text{constant}, \\
  \frac{N^\text{3D}_{\mathbf{k}\neq 0}(t_3)}{N} &\approx \frac{1}{\sqrt{V}} \cdot \frac{1}{N} \left(\frac{M U^\text{3D}_{aa} n_a}{\hbar^2}\right)^{3/2} \frac{\Lambda_a^{1/2} \cos^2\theta f^\text{3D}(\Lambda_a)}{\sqrt{2} \pi^2} &&\propto \frac{1}{\sqrt{V}}.
\end{align}
The non-condensed fraction $N_{\mathbf{k}\neq 0}/N$ thus scales differently with trap size in each of 1D, 2D and 3D.  For one-dimensional traps, the non-condensed fraction increases with system size, whereas for 2D traps it is constant.  For three-dimensional traps, the non-condensed fraction \emph{decreases} with volume.  This suggests that the best system to use will be a three-dimensional box potential, and the system size can be increased until the condensate depletion becomes negligibly small.

@@

Using a Bogoliubov analysis, however, it was shown that the coupling between different modes could be minimised in a second regime, namely the \textit{weak} trapping limit for 3D traps. This latter regime is in fact ideally compatible with large atomic numbers, and shows the prospect of extremely high squeezing.  A 3D TWA simulation using an experimentally demonstrated trapping potential yielded $15$\,dB of squeezing with $10^5$ particles. In practice, the only limit to the number of atoms in such traps would be the stability of the trap, as the BEC would become increasingly sensitive to fluctuations as the energy per particle is reduced as the volume increases.

@@

\bibitem{jeppesenET2008} Jeppesen M, Dugu{\'{e}}, Dennis G R, Johnsson M T, Figl C, Robins N P and Close J D 2008 Approaching the Heisenberg limit in an atom laser \emph{Phys. Rev.} A {\bf 77} 063618

>>>>

A precision measurement using interferometry is always performed at the maximum allowable flux in order to maximise sensitivity.  While the application of squeezing can be performed to some extent independently of the technical considerations that tend to limit flux, it is clearly most metrologically relevant when it can be achieved in the context of the maximum flux.  There are many practical limits on the usable atomic flux, typically set by the available trap parameters and loss rates.  Squeezing has already been used interferometrically on measurements using chip-based systems, which are particularly limited in their maximum atomic number \cite{Ockeloen2013}.   Outside very specific circumstances, however, atom interferometry is a sensitive tool on an absolute scale when the atomic flux can 
be large.  This paper examines the generation of atomic squeezing using the inherent Kerr-like nonlinearities of ultracold atoms, and demonstrates a trap geometry in which metrologically relevant levels of squeezing are present in trap geometries that are compatible with large atomic number.

@@

The total number squeezing in the multimode case is quantified by the normalised number variances $\mathcal{V}_i$ in internal states $|i\rangle \in \{|a\rangle, |b\rangle \}$:
\begin{equation}
\mathcal{V}_i = \langle\mathrm{Var}[\hat{N}_i] \rangle / \langle \hat{N}_i \rangle.
\label{eqMultimodeSqueezingDefinition}
\end{equation}
In the multimode case, this is calculated using the total number operator $\hat{N}_i = \sum_j \hat{N}_{i j}$, where $\hat{N}_{i j}$ is the number operator for the state with spatial mode $j$ and internal state $|i\rangle$, which is independent of the spatial mode basis.

@@

The simulations in the previous two sections were one dimensional, and show how having spatially constant atomic density and using a $\pi$-pulse will improve the resultant squeezing.  While these conclusions remain valid for simulations in two or three dimensions, adding each additional dimension increases the number of spatial modes in a given energy range.  For BECs of different dimension \textit{at the same energy per particle}, this leads to a degradation of squeezing, even when the potential (and the resulting atomic density) is constant across the bulk of the trap.  This is demonstrated in Fig.~\ref{figDimensionalSqueezingEffects}, where we show the squeezing generated in 1D, 2D and 3D BECs for two different parameter regimes. In both cases we take the BEC to be trapped in a box potential, which in the Thomas-Fermi limit leads to a constant density profile across the BEC. In Figure~\ref{figDimensionalSqueezingEffects}a we consider the case where $U_{aa}=U_{bb}\neq 0$, $U_{ab}=0$, while in Figure~\ref{figDimensionalSqueezingEffects}b we consider the case $U_{aa}\neq 0$, $U_{ab}=U_{bb}=0$. In both cases it is clear that number squeezing is best in lower dimensions, and becomes degraded or non-existent as we move to a three dimensional system; the effect is worse if more than one nonlinearity is non-zero.

@@

We wish to examine the scaling of the non-condensed fraction $N_{\mathbf{k}\neq0}/N$ with the physical size of the condensate, \textit{where the total number $N$  is kept constant}.  This is importantly different to the case above where the density is kept constant, as here the density and energy per particle will go down with volume.  Other system parameters such as the occupation of the $\ket{a}$ and $\ket{b}$ states at $t_1$ ($n_a$ and $n_b$, respectively) are also kept constant and chosen to maximise the squeezing of the equivalent two-mode model presented in Section~\ref{secTwoModeAnalytic}.  In particular, this fixes the values of $\theta$, $\phi$, and $\lambda_{ij} = \chi_{ij}\tau_\text{hold}$.  As the size of the condensate is scaled, $\hbar \chi_{ij} = U^\text{1D}_{ij}/L = U^\text{2D}_{ij}/A = U^\text{3D}_{ij}/V$ will change, but we will vary $\tau_\text{hold}$ to keep $\lambda_{ij}$ constant.  Therefore $\Lambda_a = n_a\chi_{aa}\tau_\text{hold}$ will also be constant.  With this in mind, we can write $\tau_\text{hold}$ in terms of the system size and the nonlinearity $U_{aa}$.  The non-condensed fraction can then be shown to scale with the physical size of the condensate as:
\begin{align}
  \frac{N^\text{1D}_{\mathbf{k}\neq 0}(t_3)}{N} &\approx \sqrt{L} \cdot \frac{1}{N} \sqrt{\frac{M U^\text{1D}_{aa} n_a}{2 \hbar^2 \pi^2}} \Lambda_a^{3/2} \cos^2\theta f^\text{1D}(\Lambda_a) &&\propto \sqrt{L}, \\
  \frac{N^\text{2D}_{\mathbf{k}\neq 0}(t_3)}{N} &\approx 1 \cdot \frac{1}{N} \frac{M U^\text{2D}_{aa} n_a }{4 \hbar^2} \Lambda_a \cos^2\theta f^\text{2D}(\Lambda_a) &&\propto \text{constant}, \\
  \frac{N^\text{3D}_{\mathbf{k}\neq 0}(t_3)}{N} &\approx \frac{1}{\sqrt{V}} \cdot \frac{1}{N} \left(\frac{M U^\text{3D}_{aa} n_a}{\hbar^2}\right)^{3/2} \frac{\Lambda_a^{1/2} \cos^2\theta f^\text{3D}(\Lambda_a)}{\sqrt{2} \pi^2} &&\propto \frac{1}{\sqrt{V}}.
\end{align}
The non-condensed fraction $N_{\mathbf{k}\neq 0}/N$ thus scales differently with trap size in each of 1D, 2D and 3D.  For one-dimensional traps, the non-condensed fraction increases with system size, whereas for 2D traps it is constant.  For three-dimensional traps, the non-condensed fraction \emph{decreases} with volume.  This suggests that the best system to use will be a three-dimensional box potential, and the system size can be increased until the condensate depletion becomes negligibly small.  We have therefore determined that dimensionality of the system switches from a curse to a blessing, provided we allow the density to go down as we increase the volume.  This does not limit the total number accessible, but it means that the volume must go up faster than the number to maintain high squeezing.

@@

Using a Bogoliubov analysis, however, it was shown that the coupling between different modes could be minimised in a second regime, namely the \textit{weak} trapping limit for 3D traps. This latter regime is in fact ideally compatible with large atomic numbers, and shows the prospect of extremely high squeezing.  A 3D TWA simulation using an experimentally demonstrated trapping potential yielded $15$\,dB of squeezing with $10^5$ particles. In practice, the only limit to the number of atoms in such traps would be the stability of the trap, as the BEC would become increasingly sensitive to fluctuations as the energy per particle is reduced as the volume increases.  These trap geometries would most obviously be suited for precision inertial measurements rather than in-trap measurements with goals of spatial resolution.

@@

\bibitem{jeppesenET2008} Jeppesen M, Dugu{\'{e}} J, Dennis G R, Johnsson M T, Figl C, Robins N P and Close J D 2008 Approaching the Heisenberg limit in an atom laser \emph{Phys. Rev.} A {\bf 77} 063618

\bibitem{Ockeloen2013} Ockeloen C F, Schmied R, Riedel M F, and Treutlein P, \emph{Phys. Rev. Lett.} {\bf 111} 143001 (2013).
